---
title: "Final Project 444"
author: "Donald Rogers"
date: "2022-11-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "latex")
```

```{r packages, include=FALSE}
pacman::p_load(tidyverse, arrow, sf, usmap, maps, RColorBrewer,cowplot,kableExtra)
```

```{r load-data, include=FALSE}
# Question 1 & 3 data
countyGeneration <- read_csv("./data/organised_Gen.csv")
us_states <- us_map("states")

# Question 2 data
# https://www.kaggle.com/datasets/rasyidstat/ds4g-additional-dataset
stateGen <- readxl::read_xls("./data/annual_generation_state.xls")
stateEmis <- readxl::read_xls("./data/emission_annual.xls")
```

```{r q2-wrangling, include=FALSE}
# Make a dataframe containing all emissions (except the totals) for each state and the entire US in 2018
Emis18 <- stateEmis %>%
  filter(Year == 2018) %>%
  filter(`Energy Source` != "All Sources") %>%
  select(State, `Energy Source`, `CO2
(Metric Tons)`, `SO2
(Metric Tons)`, `NOx
(Metric Tons)`) %>%
  rename("Source" = `Energy Source`)

# Combining two columns to make a left join easier with the other dataset
Emis18$combine <- paste(Emis18$State, Emis18$Source, sep = "," )

# Putting data into a single entry for each source per state.
Emis18 <- Emis18 %>%
  group_by(combine) %>%
  summarize(CO2 = sum(`CO2
(Metric Tons)`),SO2 = sum(`SO2
(Metric Tons)`),NOx = sum(`NOx
(Metric Tons)`))

# Making a dataframe containing all of the power generation in 2018 by state
Gen18 <- stateGen %>%
  filter(YEAR == 2018) %>%
  filter(`ENERGY SOURCE` != "Total") %>%
  select(STATE, `ENERGY SOURCE`, `GENERATION (Megawatthours)`) %>%
  rename("Source" = `ENERGY SOURCE`)

# Combining for a left_join
Gen18$combine <- paste(Gen18$STATE, Gen18$Source, sep = ",")

# Getting a single observation for source per state
Gen18 <- Gen18 %>%
  group_by(combine) %>%
  summarize(MWh = sum(`GENERATION (Megawatthours)`))

# Combining the two datasets together
combined18 <- left_join(Gen18, Emis18, by = "combine") %>%
  separate(combine, c("State","Source"), extra = "merge", fill = "left")

# Converting from raw numbers into tons per MWh
emissionsMWh <- combined18 %>%
  na.omit() %>%
  mutate("Tons CO2/MWh" = CO2/MWh,"Tons SO2/MWh" = SO2/MWh,"Tons NOx/MWh" = NOx/MWh,) %>%
  select(State, Source, `Tons CO2/MWh`, `Tons SO2/MWh`, `Tons NOx/MWh`)
```

```{r q3-wrangling, include=FALSE}
# Getting generation per state for 2021
state_gen <- countyGeneration %>%
  filter(YEAR == 2021) %>%
  group_by(STATE) %>%
  summarize(sum(`GENERATION (Megawatthours)`)) %>%
  rename("abbr" = "STATE",
         "gen" = "sum(`GENERATION (Megawatthours)`)") %>%
  mutate(gen = gen/1000000)

# Making dataframe for chloropleth
state_gen_map <- left_join(us_states, state_gen, by = "abbr")

# Generation per state for 2001, 2005, 2010, 2015, and 2020
gap5yr <- countyGeneration %>%
  filter(YEAR %in% c(2001,2005,2010,2015,2020)) %>%
  group_by(YEAR, STATE) %>%
  summarize(sum(`GENERATION (Megawatthours)`)) %>%
  rename("abbr" = "STATE",
         "gen" = "sum(`GENERATION (Megawatthours)`)") %>%
  mutate(gen = gen/1000000)

# Making dataframe with coordinates for map
gap5yr_map <- left_join(us_states,gap5yr, by = "abbr")

```

### Introduction

Our group decided to do a project examining the role that Green Energy plays within the US and how emissions from energy generation can affect the population surrounding them. My focus was on more generalized information, especially looking at the US from a country-wide scale. To start, I first examined how energy was generated in the US.

#### Question 1: What is the historical trend for clean energy since 2001 in the US?

First, I generated a line graph detailing the different amounts of energy producers within the US. The dataset had data from 2001-2022, but the 2022 data was incomplete, so the graphs only show up until 2021. This visualization only takes into account the total amount of facilities devoted to each resource, instead of total power generated, but provides a good idea of how our country's energy footprint has changed over the years.

```{r question-1a, message=FALSE}
# scatterplot with line of data from all years before 2022, colored by source
countyGeneration %>%
  filter(YEAR < 2022) %>%
  group_by(YEAR, `ENERGY SOURCE`) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = YEAR, y = count, group = `ENERGY SOURCE`, color=`ENERGY SOURCE`))+
  geom_line(aes(color = `ENERGY SOURCE`)) +
  geom_point(aes(color = `ENERGY SOURCE`)) +
  labs(title = "Number of Energy Producers in the US", subtitle = "(2001-2021)", y = "Number of Producers",
       x = "Year", color = "Energy Source") +
  theme_classic()
```
  
As can be seen above, energy producers based off of fossil fuels (or "Brown Energy") are still dominant, with natural gas and petroleum topping the charts, but there have been increases to the amount of green energy producers, especially since 2010. Next, let's filter out the brown energy sources so that we can see their trends better.

```{r question-1b,message=FALSE}
# Same graph but filtering out all sources of brown energy
countyGeneration %>%
  filter(YEAR < 2022) %>%
  filter(`ENERGY SOURCE` %in% c("Geothermal","Hydroelectric Conventional","Nuclear",
                                "Other Biomass", "Pumped Storage","Solar Thermal 
                                and Photovoltaic", "Wind", "Wood and Wood Derived Fuels")) %>%
  group_by(YEAR, `ENERGY SOURCE`) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = YEAR, y = count, group = `ENERGY SOURCE`, color=`ENERGY SOURCE`))+
  geom_line(aes(color = `ENERGY SOURCE`)) +
  geom_point(aes(color = `ENERGY SOURCE`)) +
  labs(title = "Number of Green Energy Producers in the US", subtitle = "(2001-2021)", 
       y = "Number of Producers", x = "Year", color = "Energy Source") +
  theme_classic()
```
  
As we see here, most of the different types of green energy have had a large increase since 2001. Solar, Thermal, and Photovoltaic have had the most growth, with Biomass and Wind following close behind. Surprisingly enough, the only form of energy generation that seems to decline is Nuclear. This is likely due to the fact that nuclear energy is seen as dangerous, although many of those concerns have been answered in recent years.




#### Question 2: What types of energy are most efficient within the United States?

Moving to possibly answer the previous question, I next examined which forms of energy were the most efficient, in terms of generating the least amount of pollutants per Megawatt hour of electricity. To provide some contest, an average American household uses about 10 MWh of power per year. The dataset also has emissions separated into three different types. They are Carbon Dioxide, Sulfur Dioxide, and Nitrogen Oxides (NO and NO2). To give a frame of reference for the amount of pollutants generated, driving a car for a year will generate about 4.5 metric tons of CO2. Sulfur Dioxide and Nitrogen Oxides are much more harmful than CO2, and are generated much less frequently than CO2 by consumers. Moving to the visualizations, I first found how many Megawatt hours of energy were generated per state in 2018 (2018 was the most recent year there was data for within the dataset), then how many metric tons of CO2, SO2, and NOx were generated per state in 2018. After joining the two datasets together, the following table was created.
  
```{r question-2a, message=FALSE}
# Table showing all states, their emissions by source, and their power generated by source
combined18 %>%
  rename("MWh Generated" = "MWh") %>%
  head(50) %>%
  kbl(caption = "Power Generation and Emissions") %>%
  kable_minimal() %>%
  add_header_above(c(" " = 3, "Gas Emissions (Metric Tons)" = 3))
```

  
The previous table showed only the first 50 observations and with over 500 observations in total, there's not much purpose to putting all of them in one table. So, since we are focusing on the entire US, let's look at the totals for the entire US to get a good idea of what energy source is the most efficient.
  
```{r question-2b, message=FALSE}
# Table filtered to just totals from US
combined18 %>%
  filter(State == "US") %>%
  rename("MWh Generated" = "MWh", "Energy Source" = "Source") %>%
  kbl(caption = "US Total Power Generation and Emissions") %>%
  kable_minimal() %>%
  kable_styling(latex_options = "hold_position") %>%
  add_header_above(c(" " = 3, "Gas Emissions (Metric Tons)" = 3))
```
  
As the table shows, there are many different types of energy that don't have any emissions recorded for those types of power plants. This is to be expected, as many of these sources don't generate any emissions since they don't require anything to be burnt to generate power. Nuclear, while appearing to generate emissions, only generates steam, since the nuclear reactions in the core of the plant are just used to heat water and don't make any gases by themselves. So, to answer the question of which energy generation is the most efficient (in terms of amount of emissions), we have our answer: Hydroelectric, Nuclear, Solar, Thermal, Photovoltaic, and Wind. I didn't include Pumped Storage, as the table shows that we actually lose power in total when using pumped storage. To see which forms of energy that create emissions are most efficient, all of the sources with no emissions were dropped for the next table.
```{r question-2c, message=FALSE}
# Further filtered to not include sources with no emissions in all three categories
combined18 %>%
  na.omit() %>%
  filter(State == "US") %>%
  rename("MWh Generated" = "MWh", "Energy Source" = "Source") %>%
  kbl(caption = "US Total Power Generation and Emissions") %>%
  kable_minimal() %>%
  kable_styling(latex_options = "HOLD_position") %>%
  add_header_above(c(" " = 3, "Gas Emissions (Metric Tons)" = 3))
```
  
As we can see from the remaining fuels, most of the emissions created in the US come from Brown Energy, such as Coal and Natural Gas, which also account for a majority of the energy created. Before doing any calculations, we can see that Natural Gas is much more efficient than Coal when it comes to emissions, as there were more MWh generated by Natural Gas sources yet less emissions in every category. Some forms of Green Energy (such as Wood, Biomass, and Geothermal) are still found in this table since they produce some forms of emissions, but not in all three categories.

```{r question-2d, message=FALSE}
# Puts CO2 calculations in table and drops any value that generates 0 tons/mwh
emissionsMWh %>%
  filter(State == "US") %>%
  select(State, Source, `Tons CO2/MWh`) %>%
  rename("Energy Source" = "Source") %>%
  filter(`Tons CO2/MWh` > 0) %>%
  arrange(`Tons CO2/MWh`) %>%
  kbl(caption = "Carbon Dioxide Generation per Megawatt Hour") %>%
  kable_minimal() %>%
  kable_styling(latex_options = "hold_position")
```
```{r question-2e, message=FALSE}
# Same as above but with SO2
emissionsMWh %>%
  filter(State == "US") %>%
  select(State, Source, `Tons SO2/MWh`) %>%
  rename("Energy Source" = "Source") %>%
  filter(`Tons SO2/MWh` > 0) %>%
  arrange(`Tons SO2/MWh`) %>%
  kbl(caption = "Sulfur Dioxide Generation per Megawatt Hour") %>%
  kable_minimal() %>%
  kable_styling(latex_options = "hold_position")
```
```{r question-2f, message=FALSE}
# Same as above but with NOx
emissionsMWh %>%
  filter(State == "US") %>%
  select(State, Source, `Tons NOx/MWh`) %>%
  rename("Energy Source" = "Source") %>%
  filter(`Tons NOx/MWh` > 0) %>%
  arrange(`Tons NOx/MWh`) %>%
  kbl(caption = "Nitrous Oxides (NO and NO2) Generation per Megawatt Hour") %>%
  kable_minimal() %>%
  kable_styling(latex_options = "hold_position")
```
  
As can be seen by these three tables, many different energy sources are more and less efficient for the different types of emissions. As mentioned before, some of the Green Energy sources only generate one or two types of emissions, meaning they won't appear on each table. As we can see from the tables, Natural Gas is the most efficient Brown Energy source that we have, as it has the least amount of SO2 and NOx emissions per MWh, and is only behind Geothermal energy on the CO2 table. Geothermal is considered to be a nonrenewable source of energy, although it does make emissions, which makes Natural Gas the most efficient Brown Energy on all three charts. When referencing the line graph under Question 1, we see that Coal and Petroleum had the most Energy Producers within the US, but when comparing them to Natural Gas, we can see that Natural Gas is anywhere from twice as efficient (Tons CO2/MWh) to 20 times as efficient (Tons SO2/MWh) as its main competitors.

#### Question 3: Which US States generate the most energy?

This question was relatively straightforward, as I wrangled one of our datasets to group the MWh generated by state in 2021 (most recent year with complete data). I put the data into a Choropleth Map, which can be seen below.
```{r question-3, message=FALSE}
# Centroid for putting total generation on choropleth
centroid <- aggregate(data = state_gen_map, cbind(x, y) ~ gen, FUN = mean)

# Choropleth of US as heat map of generation
state_gen_map %>%
  ggplot(mapping = aes(x = x, y = y, fill = gen)) +
  geom_polygon(mapping = aes(group = group)) +
  geom_text(data = centroid, mapping = aes(x = x, y = y, label =sprintf("%0.0f", gen))) +
  scale_fill_distiller(palette = "YlOrRd", trans = "reverse") +
  coord_equal() +
  theme_map() +
  labs(title = "US Power Generation by State", subtitle = "2021",
       x = element_blank(), y = element_blank(), fill = "Generation\n(TWh)")

```
  
Looking at the map, the answer to the question becomes obvious. Texas, by almost double, generates the most power in the entire US. The EIA reports that most of this huge power draw is based off of their industrial sector, accounting for over half of their total power generation. This begs the question, has Texas always been the most power hungry state? Going back to 2001, and jumping 5 years each time, the following choropleths are generated.  

```{r question-3b, message=FALSE}
# heatmap of generation wrapped by year
gap5yr_map %>%
  ggplot(mapping = aes(x = x, y = y, fill = gen)) +
  geom_polygon(mapping = aes(group = group)) +
  scale_fill_distiller(palette = "YlOrRd", trans = "reverse") +
  coord_equal() +
  theme_map() +
  labs(title = "US Power Generation by State",
       x = element_blank(), y = element_blank(), fill = "Generation\n(TWh)") +
  facet_wrap(~YEAR)
```
  
After looking at these choropleths, we can see that Texas has been far ahead of every state in terms of power generation for the last 20 years, and has only been increasing the amount of power generated in each subsequent year. Looking at the map, most states seem to be generating more power as we progress through the 21st Century, but the general ranking of highest producers seems to be consistent, with Texas at the top, and California, Florida, Illinois, and Pennsylvania being right behind. The top of the Northeast (above New York) seems to generate the least amount of power per state, with all of the states falling behind the Southeast and Midwest, but the West, save california, Arizona, and Washington doesn't seem to be much higher.

### Conclusion

Overall, we can see a few key things. First, the US seems to be building more and more Green Energy sources, but still isn't replacing a large amount of Brown Energy sources, instead supplementing them with Green Energy. Second, we can see that every Green Energy source is infinitely more efficient than any Brown Energy source, but of the different types of Brown Energy, Natural Gas causes far and away the least amount of emissions per MWh. Third, and finally, we see that Texas is the most power-hungry state in the US, and has been for the entire 21st century. We also saw that power generation has been increasing across the board in the last two decades, but all states seem to be increasing at a similar rate. Moving forward, we should look to push more and more Green Energy sources to the US so that we can start to decrease the amount of Brown Energy producers, which would also cut emissions. If we can't push Green Energy everywhere though, we should make sure to recommend Natural Gas as the next best option to reduce the amount of emissions in the US and then save our planet for generations to come.